Consistent ecosystem functional response across precipitation extremes in a sagebrush steppe

### 2017-07-26 manuscript/usses_water_ms.Rmd
addresses PBA minor comments

---
title: Consistent ecosystem functional response across precipitation extremes in a sagebrush steppe
preprint: false
author: 
  - name: Andrew T. Tredennick
    affiliation: 1
    corresponding: true
    email: atredenn@gmail.com
  - name: Andrew R. Kleinhesselink
    affiliation: 2
  - name: Bret Taylor
    affiliation: 3
  - name: Peter B. Adler
    affiliation: 1
affiliation:
  - code: 1
    address: Department of Wildland Resources and the Ecology Center, Utah State University, Logan, Utah 84322
  - code: 2
    address: Department of Ecology and Evolutionary Biology, University of California, Los Angeles, Los Angeles, California 90095
  - code: 3
    address: USSES, address...
abstract: >
  Precipitation is predicted to become more variable in the western U.S., meaning years of above and below average precipitation will become more common. Such periods of drought and deluge are major drivers of plant community dynamics and ecosystem functioning in water limited grasslands. Here we report the results of a five-year experiment where we used drought and irrigation treatments (50% decrease/increase) to see how a sagebrush steppe plant community in Idaho will respond to future climate changes. Ecosystem functional response, defined as the relationship between soil moisture and aboveground net primary productivity, was suprisingly consistent across drought, control, and irrigation treatments. Aboveground net primary productivity (ANPP) responded positively to availabile soil moisture, but the response did not vary across treatments. There was also no evidence that treatment effects grew over time. The similarity of ecosystem functional response across treatment was not due to compensatory shifts at the plant community level, where species composition among treatments was similar and remarkably stable over the five years. At least in the short-term, ecosystem functional response and community composition in this sagebrush steppe system is resistant to increases and decreases in growing season precipitation.
header-includes: >
  \usepackage{lipsum}
  \usepackage{setspace}
  \usepackage{todonotes}
  \usepackage{rotating}
  \usepackage{color, soul}
  \usepackage{sectsty}
  \usepackage{bm,mathrsfs}
bibliography: ~/Dropbox/Bibliography/usses_water.bib
output: 
  rticles::peerj_article:
    base_format: rmarkdown::pdf_document # bookdown::pdf_document2 # for using \@ref()
---

@@

@@

@@

To that end, here we report the results of a five-year precipitation manipulation experiment in a sagebrush steppe grassland.
We imposed drought and irrigation treatments (50\% decrease/increase) and measured ecosystem (ANPP) and community (species composition) responses.
We focus on how the drought and irrigation treatments affect the relationship between available soil moisture and ANPP, and if community dynamics underly the ecosystem responses.
In particular, we are interested in the consistency of the soil moisture-ANPP relationship among treatments.
Is the relationship steeper under the drought treatment, at low soil moisture?
Does the relationship saturate under the irrigation treatment, at high soil moisture?
To answer these questions we fit a random intercepts, random slopes model to test whether the regressions differed among treatments.
We also analyzed community composition over time, allowing us to place our experimental results within the framework of the HRF and our competing predictions (Fig. 1).

<!---
Chronic alterations to resource availability should cause community composition to shift over time, if the focal ecosystem is sensitive to the manipulated resource.
Species' relative abundances are expected to re-order and, eventually, some species will be lost from the local community and new species will be gained [@Smith2009; @Avolio2015].
On the heels of changes in community composition, subsequent changes in the level of ecosystem functioning may occur.

As the rate of climate change accelerates, ecologists are being pressed to forecast its impacts on ecosystem functions and services.
Making such forecasts faces two challenges.
First, future conditions are likely to be outside the historical range of variability [@Smith2011], meaning we cannot simply look to the past to predict the future.
Second, ecosystems will likely exhibit unique responses to climate change induced resource alterations [e.g., @Byrne2017], meaning we cannot simply look across space to predict the future.
These two challenges motivate the use of *in situ* experimental manipulations of resource availability [@Avolio2015].

Manipulating potentially limiting resources offers a route to understanding how ecosystems will respond to resource levels that fall outside the historical range of variability [@Knapp2017].
Chronic alterations to resource availability should cause community composition to shift over time, the focal ecosystem is sensitive to the manipulated resource.
Species' relative abundances are expected to re-order and, eventually, some species will be lost from the local community and new species will be gained [@Smith2009; @Avolio2015].
On the heels of changes in community composition, subsequent changes in the level of ecosystem functioning may occur.

@Smith2009 introduced the 'Hierarchical Response Framework' for understanding the interplay of community composition and ecosystem functioning in response to resource manipulations over time.
In the near term, ecosystem functioning such as annual net primary productivity (ANPP) will reflect the physiological responses of individual species to the manipulated resource level.
For example, ANPP may decline under simulated drought because the initial community consisted of drought-intolerant species [@Hoover2014].
Over longer time spans, ecosystem functioning may recover as new species colonize or initial species reorder in relative abundance.
For example, ANPP may initially decline, but eventually rise back to pre-treatment levels once drought-tolerant species become more abundant and compensate for drought-intolerant species [@Hoover2014].
It is also possible that ecosystem functioning shifts to a new mean state, reflective of the suite of species in the new community [@Knapp2012].

Much of the research on ecosystem and community responses to global climate change has focused on grassland systems, where soil water is typically a limiting resource.
The sensitivity of ANPP to water availability in grasslands has been characterized spatially (across sites) and temporally (within sites).
Across sites, there is a strong positive relationship between the amount of precipitation at a given site and mean ANPP [@Sala1988; @Huxman2004].
Within sites, however, the response of ANPP to annual precipitation amount is much weaker [@Huxman2004; @Hsu2012].
Furthermore, it is becoming clear that grassland ecosystem response to altered water availability is likely to be idiosyncratic, conforming to neither cross-site nor within-site ANPP-precipitation relationships [@Wilcox2016].
This is because within-site relationships do not account for species reordering under chronic resource alterations, while cross-site relationships may overestimate the ability of species compensation.
Thus, discovering which ecosystems are most sensitive to chronic alterations of water availability, and the time scales at which ecosystem and community responses occur, is critical to avoid surprises in the face of climate change.
---->

@@

## Data Collection
We estimated aboveground net primary productivity (ANPP) using a radiometer to relate ground reflectance to plant biomass [see @Byrne2011 for a review].
We recorded ground reflectance at four wavelengths, two associated with red reflectance (626 nm and 652 nm) and two associated with near-infrared reflectance (875 nm and 859 nm).
At each plot in each year, we took four readings of ground reflectances at the above wavelengths.
We also took readings in ten calibration plots adjacent to the experimental site, in which we harvested all aboveground biomass, dried it to a constant weight at 60$^{\circ}$C, and weighed it to estimate ANPP.

For each plot and year, we averaged the four readings for each wavelength and then calculated NDVI using the MODIS and AVHRR algorithms.
To convert NDVI to ANPP we regressed NDVI against the dry biomass weight from the ten calibration plots.
We fit regressions to MODIS-based NDVI and AVHRR-based NDVI for each year and retained the regression with the better fit.
Using the best regression equation for each year, we predicted ANPP (Appendix 1).

Species composition data came from two sources: yearly census maps for each plot made using a pantograph [@Hill1920] and yearly counts of annual species in each plot.
The maps record the spatial location of all individuals in the plot and the basal cover of each individual with cover greater than 1 cm.
Using those annual maps, we aggregated over individuals to calculate total basal cover for each species in each plot.
We made a large plot-treatment-year by species matrix, where columns were filled with either basal cover or density, depending on the measurement made for the particular species.
So we could analyze the different types of data together, we standardized the values in each column.
This puts all abundance values on the same scale, but comes with the limitation that all species are weighted equally.
Nonetheless, this scaling approach allows a comprehensive view of community composition dynamics through time.

## Data Analysis
Our main goal was to test whether the relationship between ANPP and growing season precipitation (hereafter, precipiation) differed among the drought, control, and irrigation treatments.
To achieve this goal, we fit a multi-level random intercept and random slope regression with log(ANPP) as the response variabile and soil moisture (volumetric water content) as the sole predictor.
We fit the model under a Bayesian framework, allowing us to test for treatment differences by comparing the posterior distributions of the treatment-level coefficients [e.g., @Tredennick2013].
Both log(ANPP) and soil moisture were standardized to have mean 0 and unit variance before fitting the model [i.e., $(x_i - \bar{x})/\sigma_x$].

Our multi-level model has three grouping levels for coefficients, representing the nested structure of the data: (i) overall coefficients, (ii) treatment coefficients, and (iii) plot coefficients.
Each subsequent level is drawn from the distribution of coefficients at the previous level.
Formally, our model is defined as follows:

@@

\begin{align}
\mu_{i(j(k(t)))} &= \beta_{0,j(k)} + \beta_{1,j(k)}x_i + \gamma_t, \\
y_{i(j(k(t)))} &\sim \text{Normal} \left(\mu_{i(j(k(t)))}, \sigma^2_{k} \right),
\end{align}

\noindent{}where $\mu_{i(j(k))}$ is the deterministic prediction from the regression model for observation *i* for plot *j* associated with treatment *k* in year *t*, $\beta_{0,j(k)}$ is the intercept for plot *j* associated with treatment *k*, $\beta_{1,j(k)}$ is the slope term for the effect of soil moisture for plot *j* associated with treatment *k*, $\gamma_t$ is the intercept offset for year *t*, and $\sigma^2_k$ is the process variance for treatment *k*.
Data include the standardized log(ANPP) observations $\left( y_{i(j(k(t)))} \right)$ and soil moisture ($x_i$).
Although we include observation subscript *i* on the *x*s, observations within a treatment-year all share the same soil moisture values.

The intercept and slope terms are modeled hierarchichally to account for the non-independence of observations across years within plots and to allow us to test the hypothesis that our treatments alter the ANPP-soil moisture relationship.
As noted above, plot-level coefficients are drawn from treatment-level coefficients, which are drawn from overall coefficients.
We also include a covariance structure among the intercept and slope at each level.
Formally, our hierarhical structure is as follows, where we drop the intercept (0) and slope (1) subscripts and instead refer to a vector of coefficients, $\boldsymbol{\beta}$:

@@

Cumulative growing season soil moisture had a positive effect on ANPP (mean of $\beta_{1}$ = `r round(betas$mean[8],2)`; 80\% BCI = `r round(betas$X10.[8],2)`, `r round(betas$X90.[8],2)`; 95\% BCI = `r round(betas$X2.5.[8],2)`, `r round(betas$X97.5.[8],2)`) (Fig. 3).
Ecosystem functional response was similar among treatments, with treatment level regressions between soil moisture and log(ANPP) having similar intercepts and slopes (Fig. 3A).
There was also no evidence that the treatment effects became more important over time because there was no directional trend in the random year effects (Fig. 4).

@@

Third, the ecosystem and community resistance to drought and deluge we observed could be a real phenomenon in this system, at least over the five year time span we observed.
The perennial plants in this cold desert ecosystem are tolerant to drought conditions [@Bazzaz1979; @Franks2011, A.R. Kleinhesselink, unpublished data], meaning a consistent ecosystem functional response at low soil moisture is perfectly reasonable.
For example, many of the perennial grasses in our focal ecosystem avoid drought stress by growing early in the growing season (A.R. Kleinhesselink, personal observation).
Likewise, the dominant shrub in our focal ecosystem, *Artemisia tripartita*, has access to water deep in the soil profile thanks to a deep root system [@Germino2014].
As for the irrigation treatments, the lack of a saturating response (e.g., a lower regression slope) suggests water availability remained a limiting factor at the levels of precipitation we were able to manipulate.

>>>>

---
title: Consistent ecosystem functional response across precipitation extremes in a sagebrush steppe
preprint: false
author: 
  - name: Andrew T. Tredennick
    affiliation: 1
    corresponding: true
    email: atredenn@gmail.com
  - name: Andrew R. Kleinhesselink
    affiliation: 2
  - name: Bret Taylor
    affiliation: 3
  - name: Peter B. Adler
    affiliation: 1
affiliation:
  - code: 1
    address: Department of Wildland Resources and the Ecology Center, Utah State University, Logan, Utah 84322
  - code: 2
    address: Department of Ecology and Evolutionary Biology, University of California, Los Angeles, Los Angeles, California 90095
  - code: 3
    address: USSES, address...
abstract: >
  Precipitation is predicted to become more variable in the western United States, meaning years of above and below average precipitation will become more common. Periods of drought and deluge are major drivers of ecosystem functioning in water limited grasslands, but how ecosystems respond to precipitation may change as the duration of above and below average periods increases. Changes in ecosystem functional response could reflect underlying changes in species composition or species reaching physiological thresholds at extreme precipitation levels. Here we report the results of a five-year experiment where we used drought and irrigation treatments (approximately 50% decrease/increase) to investigate whether ecosystem functional response remains consistent at extreme precipitation levels. We also use species-level adundance data to link ecosystem functional response to community dynamics. Ecosystem functional response, defined as the relationship between soil moisture and aboveground net primary productivity, was suprisingly consistent across drought, control, and irrigation treatments. Aboveground net primary productivity (ANPP) responded positively to availabile soil moisture, but the response was weak and did not vary across treatments. There was also no evidence that treatment effects grew over time. The similarity of ecosystem functional response across treatments was not due to compensatory shifts at the plant community level, where species composition among treatments was similar and remarkably stable over the five years. At least in the short-term, ecosystem functional response and community composition in this sagebrush steppe system is resistant to increases and decreases in growing season precipitation, perhaps due to bet-hedging strategies of constituent species.
header-includes: >
  \usepackage{setspace}
  \usepackage{todonotes}
  \usepackage{rotating}
  \usepackage{color, soul}
  \usepackage{sectsty}
  \usepackage{bm,mathrsfs}
bibliography: ~/Dropbox/Bibliography/usses_water.bib
output: 
  rticles::peerj_article:
    base_format: rmarkdown::pdf_document # bookdown::pdf_document2 # for using \@ref()
---

@@

\reversemarginpar

@@

Projected changes in precipitation regimes, associated with global climate change, will affect

@@

To that end, here we report the results of a five-year precipitation manipulation experiment in a sagebrush steppe grassland.
We imposed drought and irrigation treatments (approximately $\pm50\%$) and measured ecosystem (ANPP) and community (species composition) responses.
We focus on how the drought and irrigation treatments affect the relationship between available soil moisture and ANPP, and if community dynamics underly the ecosystem responses.
In particular, we are interested in the consistency of the soil moisture-ANPP relationship among treatments.
Is the relationship steeper under the drought treatment, at low soil moisture?
Does the relationship saturate under the irrigation treatment, at high soil moisture?
To answer these questions we fit a random intercepts, random slopes model to test whether the regressions differed among treatments.
We also analyzed community composition over time, allowing us to place our experimental results within the framework of the HRF and our competing predictions (Fig. 1).

@@

## Data Collection
We estimated aboveground net primary productivity (ANPP) using a radiometer to relate ground reflectance to plant biomass [see @Byrne2011 for a review].
We recorded ground reflectance at four wavelengths, two associated with red reflectance (626 nm and 652 nm) and two associated with near-infrared reflectance (875 nm and 859 nm).
At each plot in each year, we took four readings of ground reflectances at the above wavelengths.
We also took readings in 12 (2015), 15 (2012, 2013, 2014), or 16 (2016) calibration plots adjacent to the experimental site, in which we harvested all aboveground biomass, dried it to a constant weight at 60$^{\circ}$C, and weighed it to estimate ANPP.

For each plot and year, we averaged the four readings for each wavelength and then calculated a greenness index based on the same bands used to calculate NDVI using the MODIS and AVHRR bands for NDVI.
To convert the greenness index to ANPP we regressed the greenness index against the dry biomass weight from the ten calibration plots.
We fit regressions to MODIS-based index and AVHRR-based index for each year and retained the regression with the better fit.
Using the best regression equation for each year, we predicted ANPP (Appendix 1).

Species composition data came from two sources: yearly census maps for each plot made using a pantograph [@Hill1920] and yearly counts of annual species in each plot.
From those maps, we have data on the density of all annuals and perennials forbs, basal cover of perennial grasses, and canopy cover of shrubs.
We made a large plot-treatment-year by species matrix, where columns were filled with either basal cover or density, depending on the measurement made for the particular species.
So we could analyze the different types of data together, we standardized the values in each column.
This puts all abundance values on the same scale, meaning that common and rare species are weighted equally.
Nonetheless, if we assume that rare species will respond more than common ones, then this approach is anti-conservative for detecting community change.
This means that our approach is biased toward detecting compositional changes.\todo{Did I say this right?}

## Data Analysis
Our main goal was to test whether the relationship between ANPP and soil moisture differed among the drought, control, and irrigation treatments.
Based on our own observations and previous work at our study site [@Blaisdell1958; @Dalgleish2011; @Adler2012], we chose to use cumulative volumetric water content from March through June as our metric of soil moisture (hereafter referred to simply as `soil moisture').
To achieve this goal, we fit a multi-level random intercept and random slope regression with log(ANPP) as the response variabile and soil moisture as the sole predictor.
We log-transformed ANPP to account for heteroscedasticity.
We fit the model under a Bayesian framework, allowing us to test for treatment differences by comparing the posterior distributions of the treatment-level coefficients [e.g., @Tredennick2013].
Both log(ANPP) and soil moisture were standardized to have mean 0 and unit variance before fitting the model [i.e., $(x_i - \bar{x})/\sigma_x$].

Our model is defined as follows:

@@

\begin{align}
\mu_{j(k(t))} &= \beta_{0,j(k)} + \beta_{1,j(k)}x_{k(t)} + \gamma_t, \\
y_{j(k(t))} &\sim \text{Normal} \left(\mu_{j(k(t))}, \sigma^2_{k} \right),
\end{align}

\noindent{}where $\mu_{j(k(t))}$ is the deterministic prediction from the regression model for plot *j* associated with treatment *k* in year *t*, $\beta_{0,j(k)}$ is the intercept for plot *j* associated with treatment *k*, $\beta_{1,j(k)}$ is the slope term for the effect of soil moisture for plot *j* associated with treatment *k*, $\gamma_t$ is the intercept offset for year *t*, and $\sigma^2_k$ is the process variance for treatment *k*.
Data include the standardized log(ANPP) observations $\left( y_{j(k(t))} \right)$ and soil moisture ($x_{k(t)}$).

The intercept and slope terms are modeled hierarchichally to account for the non-independence of observations across years within plots and to allow us to test the hypothesis that our treatments alter the ANPP-soil moisture relationship.
Thus, our multi-level model has three grouping levels for coefficients, representing the nested structure of the data: (i) overall coefficients, (ii) treatment coefficients, and (iii) plot coefficients.
Each subsequent level is drawn from the distribution of coefficients at the previous level, meaning plot-level coefficients are drawn from treatment-level coefficients, which are drawn from overall coefficients.
We also include a covariance structure among the intercept and slope at each level.
Formally, our hierarhical structure is as follows, where we drop the intercept (0) and slope (1) subscripts and instead refer to a vector of coefficients, $\boldsymbol{\beta}$:

@@

Cumulative March-June soil moisture had a weak positive effect on ANPP (mean of $\beta_{1}$ = `r round(betas$mean[8],2)`), but the effect is associated with high uncertainty, with credible intervals that overlap zerp (80\% BCI = `r round(betas$X10.[8],2)`, `r round(betas$X90.[8],2)`; 95\% BCI = `r round(betas$X2.5.[8],2)`, `r round(betas$X97.5.[8],2)`) (Fig. 3).
Ecosystem functional response was similar among treatments, with treatment level regressions between soil moisture and log(ANPP) having similar intercepts and slopes (Fig. 3A).
There was also no evidence that the treatment effects became more important over time because there was no directional trend in the random year effects (Fig. 4).

@@

Third, the ecosystem and community resistance to drought and deluge we observed could be a real phenomenon in this system, at least over the five year time span we observed.
The perennial plants in this cold desert ecosystem are tolerant to drought conditions [@Bazzaz1979; @Franks2011, A.R. Kleinhesselink, unpublished data], meaning a consistent ecosystem functional response at low soil moisture is perfectly reasonable.
For example, many of the perennial grasses in our focal ecosystem avoid drought stress by growing early in the growing season [@Blaisdell1958, A.R. Kleinhesselink, personal observation].
Likewise, the dominant shrub in our focal ecosystem, *Artemisia tripartita*, has access to water deep in the soil profile thanks to a deep root system (CITATION).
As for the irrigation treatments, the lack of a saturating response (e.g., a lower regression slope) suggests water availability remained a limiting factor at the levels of precipitation we were able to manipulate.
